---
title: "Beta part"
author: "Colin J. Brislawn"
date: "February 10, 2025"
output:
  html_document:
    toc: true
---

```{r setup0, include=T}
library("tidyverse")
library("DT")

library("vegan")
library("betapart")
library("broom")

library("ggpubr")

knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

theme_set(theme_bw())
theme_update(
  strip.background = element_blank(),
  axis.text.x = element_text(angle = -20, hjust = 0)
)
```

```{r lint, include=F, eval=F}
library(styler)
style_dir(strict = TRUE)

library(lintr)
lintr::lint("betapart.Rmd",
  linters = linters_with_defaults(
    # line_length_linter = line_length_linter(120),
    object_name_linter = NULL,
    commented_code_linter = NULL
  )
)
```

## Import binary feature table

```{r betapart}
# Feature table binary
ftab_b <- read_tsv("feature_table_binary.tsv", skip = 1, col_names = T) |>
  column_to_rownames(var = "#OTU ID") |>
  t()
# Transpose because taxa_are_rows in Qiime2 and betapart wants a
# "data frame, where rows are sites and columns are species"

stopifnot(
  "values greater than 1 detected. This table must be binary!" = max(ftab_b) == 1
  # There's more to check, but this would catch if the user forgot to normalize
)

ftab_b_bp <- beta.pair(ftab_b, index.family = "jaccard")
ftab_b_bp |> glimpse()
```

## Matrix to tibble

```{r bp_tidy}
bp_tidy <- function(bp_list) {
  a <- bp_list$beta.jac |>
    tidy(diagonal = F, upper = F) |>
    mutate(Jaccard = distance, distance = NULL)
  b <- bp_list$beta.jne |>
    tidy(diagonal = F, upper = F) |>
    mutate(Nestedness = distance, distance = NULL)
  c <- bp_list$beta.jtu |>
    tidy(diagonal = F, upper = F) |>
    mutate(Turnover = distance, distance = NULL)

  a |>
    left_join(b, by = join_by(item1, item2)) |>
    left_join(c, by = join_by(item1, item2)) |>
    mutate(
      item1 = as.character(item1),
      item2 = as.character(item2)
    )
}

ftab_b_bpt <- ftab_b_bp |> bp_tidy()
ftab_b_bpt |> head()
```

## Merge in meta

Because our distances are now in a list, we have to track metadata for _pairs_ of samples, which is tricky!

```{r meta}
meta_all <- read_tsv("sample-metadata.tsv", comment = "#")

# Filter to keep samples that are in our feature table
meta <- meta_all |> filter(SampleID %in% rownames(ftab_b))
# meta |> glimpse()
```

```{r bp_dispersion}
summarize_within_group <- function(meta, bp_tidy, group_col) {
  meta |>
    inner_join(meta, by = group_col, relationship = "many-to-many") |>
    mutate(item1 = SampleID.x, item2 = SampleID.y) |>
    select(item1, item2, {{ group_col }}) |>
    filter(item1 != item2) |>
    inner_join(bp_tidy, by = join_by(item1, item2))
}
```

```{r bp_dispersion_plot_sample}
meta |>
  summarize_within_group(ftab_b_bpt, "autoFmtGroup") |>
  pivot_longer(c(Jaccard, Nestedness, Turnover), names_to = "Part", values_to = "value") |>
  ggboxplot("item1", "value", color = "Part") +
  labs(
    title = "Per-sample Jaccard, split by nestedness and turnover",
    x = "Sample",
    y = "Pairwise distance to all \nother samples within this group"
  ) +
  theme(axis.text.x = element_blank()) +
  facet_grid(Part ~ autoFmtGroup, space = "free", scales = "free")
```

```{r bp_dispersion_plot_groups, fig.height = 4, fig.width = 9}
bp_boxplot <- function(df) {
  group_col_name <- colnames(df)[3]
  over_two_groups <- length(unique(df[[group_col_name]])) > 2

  df |>
    pivot_longer(c(Jaccard, Nestedness, Turnover), names_to = "Part", values_to = "value") |>
    ggboxplot(
      group_col_name, "value",
      color = "Part",
      ylab = "Pairwise distance to all other \nsamples within this group",
      ylim = c(0, 1.14),
      yticks.by = 0.2,
      ggtheme = theme_pubclean()
    ) +
    facet_grid(~Part, space = "free", scales = "free") +
    {
      if (over_two_groups) {
        suppressWarnings(stat_compare_means())
      } else {
        suppressWarnings(stat_compare_means(method = "t.test"))
      }
    } +
    theme(
      strip.background = element_blank(),
      axis.text.x = if (over_two_groups) element_text(angle = 20, hjust = 1)
    )
}

meta |> summarize_within_group(ftab_b_bpt, "autoFmtGroup") |> bp_boxplot()

meta |> summarize_within_group(ftab_b_bpt, "Disease") |> bp_boxplot()

meta |> summarize_within_group(ftab_b_bpt, "DayRelativeToNearestHCT") |> bp_boxplot()

meta |> summarize_within_group(ftab_b_bpt, "PatientID") |> bp_boxplot()

```

```{r, fig.height = 8, fig.width = 8, warning=F}
meta |>
  mutate(Patient_Diease = paste(PatientID, Disease, sep = "_")) |>
  summarize_within_group(ftab_b_bpt, "Patient_Diease") |>
  separate_wider_delim(Patient_Diease, "_", names = c("Patient", "Disease")) |>
  bp_boxplot() +
  facet_wrap(~Disease+Part, scales = "free", ncol = 3)

```

```{r bp_dispersion_plot_stacked, include = F, eval = F}
meta |>
  summarize_within_group(ftab_b_bpt, "autoFmtGroup") |>
  pivot_longer(c(Nestedness, Turnover), names_to = "Parts", values_to = "value") |>
  head()
# see, we can stack these two to get total Jaccard!
```

```{r bp_dispersion_plot_grid, include=F, eval=F}
meta |>
  summarize_within_group(ftab_b_bpt, "autoFmtGroup") |>
  # pivot_longer(c(Nestedness, Turnover), names_to = "Parts", values_to = "value") |>
  pivot_longer(c(Jaccard, Nestedness, Turnover), names_to = "Part", values_to = "value") |>
  filter(Part == "Jaccard") |>
  ggplot(aes(x = item1, y = value, fill = Part)) +
  facet_grid(~autoFmtGroup, space = "free", scales = "free") +
  geom_bar(stat = "identity")
```

```{r, include=F, eval=F}
rmarkdown::render("betapart.Rmd")
# To clear cache
unlink("betapart_cache/")
```

```{r bp_betweeness, include=F, eval=F}
# Between groups is harder to think about and graph
# Let's save this for later
ftab_b_bpt_autoFmtGroup <- ftab_b_bpt |>
  left_join(
    meta |> mutate(item1 = SampleID, autoFmtGroup1 = autoFmtGroup, .keep = "none")
  ) |>
  left_join(
    meta |> mutate(item2 = SampleID, autoFmtGroup2 = autoFmtGroup, .keep = "none")
  ) |>
  # arrange(autoFmtGroup1, item1) |>
  arrange(autoFmtGroup1, item1, autoFmtGroup2, item2) |> # Sort the data
  mutate(
    autoFmtGroup1 = fct_rev(autoFmtGroup1), # Reverse facet order
    item1 = fct_inorder(item1), # Use current order as factor levels for item1
    item2 = fct_inorder(item2) # Use current order as factor levels for item2
  )

# mutate(
#   autoFmtGroup1 = fct_rev(autoFmtGroup1),
#   item1 = fct_reorder(item1, autoFmtGroup1),
#   item2 = fct_reorder(item2, autoFmtGroup2)
#   )

ftab_b_bpt_autoFmtGroup |>
  ggplot(aes(item1, item2, fill = Jaccard)) +
  geom_raster() +
  facet_grid(autoFmtGroup1 ~ autoFmtGroup2, space = "free", scales = "free")
```


## Unused

### Test for triangle inequality to confirm true distances

```{r testdist, include=F, eval=F}
if (
  any(is.na(ftab_b_bp$beta.jac))
) {
  stop("NAs in the matrix")
}
if (
  any(ftab_b_bp$beta.jac < 0)
) {
  stop("Negative distances produced!")
}

# Check nestedness
dist_matrix <- as.matrix(ftab_b_bp$beta.jne) # Convert to a regular matrix

if (any(is.na(dist_matrix))) stop("NAs in the matrix")
if (any(dist_matrix < 0)) stop("Negative distances produced!")

# 1. Ensure Symmetry
if (!isSymmetric(dist_matrix)) stop("Not symmetric")

# 2. Ensure Diagonal is Zero
if (any(diag(dist_matrix) != 0)) stop("The diagonal is NOT all zeros!\n")

# 3. Triangle Inequality (Most computationally intensive)
triangle_check <- function(dist_matrix) {
  n <- nrow(dist_matrix)
  for (i in 1:n) {
    for (j in 1:n) {
      for (k in 1:n) {
        if (dist_matrix[i, j] + dist_matrix[j, k] >= dist_matrix[i, k]) {
          # All good!
        } else { # Violation
          cat("Triangle inequality violated at: [", i, ", ", j, ", ", k, "]\n", sep = "")
          cat(dist_matrix[i, j], " + ", dist_matrix[j, k], " <= ", dist_matrix[i, k], "\n", sep = "")
          cat(dist_matrix[i, j] + dist_matrix[j, k], " <= ", dist_matrix[i, k], "\n", sep = "")
          return(FALSE) # Stop at the first violation
        }
      }
    }
  }
  return(TRUE) # No violations found
}
# ftab_b_bp$beta.jac |> as.matrix() |> triangle_check() # passes!
ftab_b_bp$beta.jne |>
  as.matrix() |>
  triangle_check()
ftab_b_bp$beta.jtu |>
  as.matrix() |>
  triangle_check()
```

### Adonis test

Let's use the `vegan::adonis` test to see how much variation of nestedness and
turnover can be attributed to a metadata variable.

Note that while Jaccard distances of nestedness and turnover add up to
total Jaccard distance, these R2 values will not sum up.

Rather, this qualitative measurement shows what correlates best.

```{r adonis2}
meta_all <- read_tsv("sample-metadata.tsv", comment = "#")

# Filter to keep samples that are in our feature table
meta <- meta_all |> filter(SampleID %in% rownames(ftab_b))
meta |> glimpse()

adonis2_on_bplist <- function(formula_rhs, meta, bp_list) {
  rbind(
    adonis2(
      as.formula(paste("bp_list$beta.jac ~", formula_rhs)),
      data = meta, permutations = 999
    ) |>
      rownames_to_column() |>
      tibble(test = formula_rhs, dist = "Jaccard Total")
    # Can use these with ADONIS2 because they are not true distances. They don't satisfy the triangle inequality.
    # adonis2(as.formula(paste("bp_list$beta.jtu ~", formula_rhs)),
    #   data = meta, permutations = 999
    #   ) |> rownames_to_column() |> tibble(test = formula_rhs, dist = "Jaccard Turnover"),
    # adonis2(as.formula(paste("bp_list$beta.jne ~", formula_rhs)),
    #   data = meta, permutations = 999
    #   ) |> rownames_to_column() |> tibble(test = formula_rhs, dist = "Jaccard Nestedness")
  ) |>
    filter(rowname == "Model") |>
    select(-rowname) |>
    relocate(test, dist)
}

# Example Usage
ftab_b_bp_adonis <- rbind(
  adonis2_on_bplist("PatientID", meta, ftab_b_bp),
  adonis2_on_bplist("Disease", meta, ftab_b_bp),
  adonis2_on_bplist("Timepoint", meta, ftab_b_bp),
  adonis2_on_bplist("Disease + Timepoint", meta, ftab_b_bp),
  adonis2_on_bplist("Timepoint + Disease", meta, ftab_b_bp),
  adonis2_on_bplist("autoFmtGroup", meta, ftab_b_bp)
)
```

```{r adonis2plots}
ftab_b_bp_adonis |> datatable(rownames = FALSE)

ftab_b_bp_adonis |> ggplot(aes(x = test, y = R2)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    # title = "Variance Explained by metadata variables",
    y = "ADONIS R² value (fraction of variation explained)"
  )
```
